---
title: "EDS241: Assignment 3"
author: "Desik Somasundaram"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 
  
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}

# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)


# load packages
packages=c("stargazer", "here", "tidyverse","stringr", "janitor", 
           "cowplot", "ggplot2", "tinytex", "datasets", "tibble", "readxl","estimatr", "car", "huxtable")

for (i in packages) {
  if (require(i,character.only=TRUE)==FALSE) {
    install.packages(i,repos='http://cran.us.r-project.org')
  }
  else {
    require(i,character.only=TRUE)
  }
}

#devtools::install_github('rstudio/rmarkdown')
options(scipen=10) # not scientific notation


```

\noindent This exercise asks you to implement some of the techniques presented in Lectures 6-7. The
goal is to estimate the causal effect of maternal smoking during pregnancy on infant birth weight
using the treatment ignorability assumptions. The data are taken from the National Natality
Detail Files, and the extract “SMOKING_EDS241.csv”' is a random sample of all births in
Pennsylvania during 1989-1991. Each observation is a mother-infant pair. 
The key variables are:
The outcome and treatment variables are:
birthwgt=birth weight of infant in grams
tobacco=indicator for maternal smoking
The control variables are:
mage (mother's age), meduc (mother's education), mblack (=1 if mother black), alcohol (=1 if
consumed alcohol during pregnancy), first (=1 if first child), diabete (=1 if mother diabetic),
anemia (=1 if mother anemic)

# Clean data

\noindent The following code loads and cleans the data.

```{r , include=TRUE}

# Load data
smokingdata <- read_csv("SMOKING_EDS241.csv")
# Clean data
smokingdata <-janitor::clean_names(smokingdata)
```
# Unadjusted mean difference
\noindent (a) What is the unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers? Under what assumption does this correspond to the average treatment effect
of maternal smoking during pregnancy on infant birth weight? Provide some simple empirical
evidence for or against this assumption. 

```{r , include=TRUE}
# smoking mothers mean
mu_nonsmoker = smokingdata %>% 
  filter(tobacco == 0) %>% 
  summarize(mean(birthwgt))

# non-smoking mothers mean
mu_smoker = smokingdata %>% 
  filter(tobacco == 1) %>% 
  summarize(mean(birthwgt))

# calculate mean  diff
mean_diff = as.numeric(mu_nonsmoker - mu_smoker)

# linear regression of choice covariate
model_1 <- lm_robust(meduc ~ tobacco, data = smokingdata)
huxreg(model_1)
```

**The unadjusted mean difference in birth weight of infants `r round(mean_diff, 3)` grams.** Statistically different from zero. Under the "treatment ignorability assumption", this corresponds with the average treatment effect(ATE) of maternal smoking during pregnancy on infant birth weight. Assumption of “treatment ignorability” conditional on pre-treatment characteristics that allow us to assume that smoking mothers and nonsmoking mothers are good counterfactuals. The assumption of common support ensures that there is sufficient overlap in the characteristics of smoking mothers and nonsmoking mothers to find adequate matches so this aspect would need further analysis of the data. It's important to note that the treatment of smoking is not randomly assigned either. The treatment ignorability assumption says that conditional on observable covariates, the assignment to the treatment is independent of the outcome of infant birth weight. Observational regression bias can arise if the smoking mother and nonsmoking mothers are inherently different in a way which would affect their infant birth weight. For example, smoking mothers could be less health conscious which has a negative impact on their infant birth weights aside from the effect of smoking itself. The regression of tobacco one mother's education yields a statistically significant relationship which questions the validity of the assumption. There is omitted variable bias shown by the model_1 regression which prevents us from being able to interpret the unadjusted mean difference as a causal effect. Unconditional treatment ignorability is not holding true.

\newpage

# Introducing covariates
\noindent Assume that maternal smoking is randomly assigned conditional on the observable
covariates listed above. Estimate the effect of maternal smoking on birth weight using a linear
regression. Report the estimated coefficient on tobacco and its standard error.


```{r , include=TRUE}
model_2 <- lm_robust(formula =  birthwgt ~ tobacco + mage + meduc + as.factor(anemia) 
                     + as.factor(diabete) + as.factor(alcohol) + 
                       as.factor(mblack) + as.factor(first), data=smokingdata)

model_2_ht <- huxreg(model_2)
restack_across(model_2_ht,13)
```

\noindent Table 1 shows the **estimated coefficients from the linear regression of effect of maternal smoking on birth weight. `r round(model_2$coefficients[2],3)` is the estimated coefficient on tobacco with `r round(model_2$std.error[2],3)` being the standard error**.


# Exact matching
\noindent Use the exact matching estimator to estimate the effect of maternal smoking on birth weight.
For simplicity, consider the following covariates in your matching estimator: create a 0-1
indicator for mother's age (=1 if mage>=34), and a 0-1 indicator for mother's education (1 if
meduc>=16), mother's race (mblack), and alcohol consumption indicator (alcohol). These 4
covariates will create 2*2*2*2 = 16 cells. Report the estimated average treatment effect of
smoking on birthweight using the exact matching estimator and its linear regression analogue. 

```{r , include=TRUE}
# create indicator based on mother's age (=1 if mage>=34)
smokingdata<- smokingdata %>% 
              mutate(age_indicator = case_when(
                      mage < 34 ~ 0,
                      mage >= 34 ~ 1))

# create indicator based on mother's education (1 if meduc>=16)
smokingdata<- smokingdata %>% 
              mutate(educ_indicator = case_when(
                      meduc < 16 ~ 0,
                      meduc >= 16 ~ 1))

# create group variable to capture all interactions
smokingdata<- smokingdata %>% 
              mutate(g = paste0(age_indicator,educ_indicator,mblack,alcohol))

# regression on including tobacco and the 4 grouped indicators (mother age, mother education, mother race and alcohol)
model_3 <- lm_robust(birthwgt ~ tobacco +  as.factor(g), data= smokingdata)
```



```{r , include=TRUE}
# exact matching table
TIA_table <- smokingdata %>%
  group_by(g,tobacco)%>%
  summarise(n_obs = n(),
            birthwgt_mean= mean(birthwgt, na.rm = T))%>% #Calculate number of observations and Y mean by X by treatment cells
  gather(variables, values, n_obs:birthwgt_mean)%>% #Reshape data
  mutate(variables = paste0(variables,"_",tobacco, sep=""))%>% #Combine the treatment and variables for reshaping
  pivot_wider(id_cols = g, names_from = variables,values_from = values)%>% #Reshape data by treatment and X cell
  ungroup()%>%  #Ungroup from X values
  mutate(birthwgt_mean_diff = birthwgt_mean_1 - birthwgt_mean_0, #calculate Y_diff
         w_ATE = (n_obs_0+n_obs_1)/(sum(n_obs_0)+sum(n_obs_1)),
         w_ATT = n_obs_1/sum(n_obs_1))%>% #calculate weights
  mutate_if(is.numeric, round, 2) #Round data


stargazer(TIA_table, type= "text", summary = FALSE, digits = 2)

# Multivariate matching estimates of ATE
ATE=sum((TIA_table$w_ATE)*(TIA_table$birthwgt_mean_diff))
ATE

```
\noindent  Table 2 shows the exact matching table used to estimate ATE.

**The exact matching estimator estimates the average treatment effect(ATE) of smoking on birthweight as `r round(ATE, 3)`. The linear analogue estimates the average treatment effect(ATE) of smoking on birthweight as `r round(model_3$coefficients[2],3)`.**

# Propensity Score
\noindent Estimate the propensity score for maternal smoking using a logit estimator and based on the
following specification: mother’s age, mother’s age squared, mother’s education, and indicators
for mother’s race, and alcohol consumption. 



```{r , include=TRUE}
# add mother's age variable squared
smokingdata <- smokingdata %>% 
                mutate(mage_squared = mage * mage)

model_4 <- glm(tobacco ~ mage + mage_squared + meduc + as.factor(mblack) + as.factor(alcohol), family = binomial(), data = smokingdata)
summary(model_4)
EPS <- predict(model_4, type = "response") # estimated propensity score (EPS)
PS_weighted <- (smokingdata$tobacco / EPS) + ((1 - smokingdata$tobacco)/(1 - EPS)) # weight EPS

```

\newpage
\noindent Use the propensity score weighted regression (WLS) to estimate the effect of maternal
smoking on birth weight (Lecture 7, slide 12).


```{r , include=TRUE}
# add propensity scores and weighted propensity scores to data as columns
smokingdata <- smokingdata %>% 
  mutate("EPS" = EPS, "PS_weighted" = PS_weighted)


model_5 <- lm_robust(birthwgt ~ tobacco + mage + mage_squared + meduc + 
                     as.factor(mblack) + as.factor(alcohol), data=smokingdata, weights=PS_weighted)

wls_ATE <- model_5$coefficients[2]

```
**The weighted OLS regression estimates the average treatment effect(ATE) of smoking on birthweight as `r round(wls_ATE, 3)`**

\newpage
## Appendix

```{r}
model_3_ht <- huxreg(model_3)
restack_across(model_3_ht,20)
```

\noindent Table A1 shows the estimated coefficients from the linear regression of effect of maternal smoking on birth weight with the inclusion of the following covariates: mother age, mother education, mother race and alcohol.


```{r}
model_5_ht <- huxreg(model_5)
restack_across(model_5_ht,13)
```
\noindent Table A2 shows the estimated coefficients from weighted OLS estimating the effect of maternal smoking on birth weight with the inclusion of the following covariates: mother age, mother age squared, mother education, mother race and alcohol.

